Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2023-02-23 analysis/memo_TAS_reviewers.Rmd
fixed minor errors, referred to paper's citation to Viele2014

4. Absent some very strong assumptions, there is no generally justifiable hyper prior for $\alpha$ in this context of strike bias and _Batson_ challenges. This is because the proper weight to place on historical data depends on what causes incompatibility between the historical data and the current data. The plausible reasons for non-random missingness and such confounding, however, likely vary considerably depending on strike procedure, courtroom practices, prosecutor charging practices, among other things. As we now note in the Discussion section of the paper, the literature on the power prior discusses information criteria for guide values of $\alpha$ as well as treating $\alpha$ as random and assigning a prior to it. However, that literature is unclear as to what implies for _Batson_ and peremptory strikes, for at least three reasons.  First, that literature has mostly focused on clinical trials and models whose properties are well known (e.g., the linear model), as distinct from the models of strike procedure discussed here.  Second, our approach does not attempt to maximize coverage in the traditional sense, but aims only to "detect" bias by determining whether zero falls outside the chosen credible interval for the posterior distribution of the bias parameter. Third, whereas some papers treat incompatibility as itself a parameter [e.g., @Ollier2020], the precise approaches rely on strong assumptions that do not obviously make sense in our context.  Accordingly, we believe the most conservative choice (for now) is to keep $\alpha$ a fixed value and check for sensitivity of results to the choice of $\alpha$ value. Proceeding otherwise would require the work of a completely new paper.

@@

Second, as the phrase "should be reprimanded in the future (in some cases for no reason)" implies, the apparent worry is that our approach will cause judges to infer, in any particular case, that illegal strike bias is more likely than not (i.e., find _Batson_ violations) when in fact it was less likely than not. This is not the case. Our approach does not and cannot automate findings of _Batson_ violations. See Item 1, above.

@@

Fourth, the comment is overinclusive. Most work on the use of historical strike data (a) presupposes that _Batson_ law permits prior strikes as relevant evidence of bias in the current trial; presupposes that any statistical approach therein cannot automatically determine _Batson_ violations alone; and (c) adjusts in some way for incompatibility between historical strike data and strikes from the current trial. As an example, we now point in the paper's Introduction to @Gastwirth2014. Our approach has a distinctive benefit of making incompatibility assumptions transparent via choosing a fixed value for the power prior's $\alpha$.

@@

We agree that, given high incompatibility, (1) the bias "detection" rate can be as high as 100% when current bias is low ($b_curr$); and (2) bias detection can be low where current bias is high but bias in historical trials is low. The tradeoff is that, given high compatibility, historical strike data increases power that is otherwise low due the limited number of strikes per trial. The advantage of using the power prior: We can adjust the weight of the historical strike data (and therefore type I error) by adjusting $\alpha$.

@@

We appreciate this suggestion but we are worried that this would unduly increase the paper's length, particular given that absence of a closed form solution for the posterior distribution even in a toy example. Because of this concern, we have already moved other parts of the paper to the (online) Appendix. We intend the reader to gain understanding by following along using our prototype software application.

>>>>

4. Absent some very strong assumptions, there is no generally justifiable hyper prior for $\alpha$ in this context of strike bias and _Batson_ challenges. This is because the proper weight to place on historical data depends on what causes incompatibility between the historical data and the current data. The plausible reasons for non-random missingness and such confounding, however, likely vary considerably depending on strike procedure, courtroom practices, prosecutor charging practices, among other things. As we now note in the Discussion section of the paper, the literature on the power prior discusses information criteria for guide values of $\alpha$ as well as treating $\alpha$ as random and assigning a prior to it. However, that literature is unclear as to what implies for _Batson_ and peremptory strikes, for at least three reasons.  First, unlike models of strike procedure discussed here, that literature has mostly focused on clinical trials and models whose properties are well known (e.g., the linear model). In the clinical trial setting, it is easier to measure trial-level (and patient-level) characteristics that may be responsible for incompatibility. Even in these settings, selecting a prior for $\alpha$ is not straightforward. @Neelon2010 warn that using random power parameters (in their case, with a beta prior distribution) tends to attenuate (more down-weighting) the influence of the historical data, and so one might have to use highly informative priors on the power parameter. These authors suggest that practitioners consider whether "it is more appropriate to assign $\alpha$ a fixed value based on expert opinion about the relevance of the historical data to the current analysis". More recent versions of the power prior are beginning to address this issue (e.g., using "commensurate power priors" [@Hobbs2011]). Second, our approach does not attempt to maximize coverage in the traditional sense, but aims only to "detect" bias by determining whether zero falls outside the chosen credible interval for the posterior distribution of the bias parameter. Third, whereas some papers treat incompatibility as itself a parameter [e.g., @Ollier2020], the precise approaches rely on strong assumptions that do not obviously make sense in our context.  Accordingly, we believe the most conservative choice (for now) is to keep $\alpha$ a fixed value and check for sensitivity of results to the choice of $\alpha$ value. Proceeding otherwise would require the work of a completely new paper.

@@

Second, as the phrase "should be reprimanded in the future (in some cases for no reason)" implies, the apparent worry is that our approach will cause judges to infer, in any particular case, that illegal strike bias is more likely than not (i.e., find _Batson_ violations) when in fact it was less likely than not. This is not the case. Our approach does not and cannot automate findings of _Batson_ violations. See Item 1(1) above.

@@

Fourth, the comment is overinclusive. Most work on the use of historical strike data (a) presupposes that _Batson_ law permits prior strikes as relevant evidence of bias in the current trial; (b) presupposes that any statistical approach therein cannot automatically determine _Batson_ violations alone; and (c) adjusts in some way for incompatibility between historical strike data and strikes from the current trial. As an example, we now point in the paper's Introduction to @Gastwirth2014. Our approach has a distinctive benefit of making incompatibility assumptions transparent via choosing a fixed value for the power prior's $\alpha$.

@@

We agree that, given high incompatibility, (1) the bias "detection" rate can be as high as 100% when current bias is low ($b_{curr}$); and (2) bias detection can be low where current bias is high but bias in historical trials is low. The tradeoff is that, given high compatibility, historical strike data increases power that is otherwise low due the limited number of strikes per trial. The advantage of using the power prior: We can adjust the weight of the historical strike data (and therefore type I error) by adjusting $\alpha$.

@@

We appreciate this suggestion but we are worried that this would unduly increase the paper's length, particularly given the lack of a closed form solution for the posterior distribution even in a toy example. Because of this concern, we have already moved other parts of the paper to the (online) Appendix. We intend the reader to gain understanding by following along using our prototype software application. We also now refer the reader to a gentle introduction to the power prior in @Viele2014.
